---
title: "Using readwritesqlite"
author: "Joe Thorley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using readwritesqlite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`readwritesqlite` is an R package to enhance reading and writing to SQLite databases.

## Starting

The first task after loading `readwritesqlite` is to create an object of class SQLiteConnection.
Below we create one in memory although in general the user will want to specify a path.
```{r}
library(readwritesqlite)
conn <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
```

## Options

The readwritesqlite package has one option.

- `rws.silent` a flag which specifies whether package specific messages and warnings should be suppressed.

## Writing Data

Individual data frames or environments or named lists of data frames can be written to a connection using `rws_write_sqlite()`.

In the case of a data frame the default table name is the name of the object.
```{r}
rws_write_sqlite(rws_data, exists = FALSE, conn = conn)
rws_list_tables(conn)
```
The fact that `rws_write_sqlite()` accepts environments means that the user can easily
write all the data frames in the current environment to a SQLiteConnection.
```{r}
a_table <- rws_data[c("date", "logical")]
another_table <- rws_data[c("factor", "ordered")]
not_a_table <- 1
rws_write_sqlite(environment(), exists = FALSE, complete = FALSE, conn = conn)
rws_list_tables(conn)
```
Objects which are not data frames are silently filtered from enviromnents but cause an error with lists.

### Exists

By default the `exists` argument is `TRUE` which means only existing tables can be writted to.

If the user 

### Foreign Keys

The `rws_write_sqlite()` function ensures that writing the new data frame(s) to the database does not violate foreign keys.
If any data does the database is left unchanged.

### Commiting Data

By default, if no check or key violations occur the data frame(s) are written to the database.
If, however, the user wishes to simply confirm that data could be written to a database without actually making any changes then they
can call `rws_write_sqlite()` with `commit = FALSE`.

### Deleting Data (and Meta Data)

Meta data is recorded if the user uses `rws_write_sqlite()` to write data to an empty table.
In order to the change the meta data the user should read the existing data from the table 
(using `rws_read_sqlite()`), modify it accordingly and then re-write it using 
`rws_write_sqlite()` with `delete = TRUE`. 
The only exception is for factors and ordered factors.
In this case the user can pass data with additional or rearranged factor or ordered factor levels without deleting the existing data.

### Complete Data

If the user sets `complete = TRUE` when writing enviroments or named lists of data frames
then the function issues an informative error message if any of data frames have duplicated names or some of the existing tables are not represented.
If it is also the case that `exists = TRUE` then extra data frames are ignored and an informative warning message is issued.
The combination of `exists = TRUE` and `complete = TRUE` in conjunction with an environment of data frames
is really useful for ensuring all tables are populated while automatically removing extraneous objects and data frames.

## Reading Data

Data can be read using `rws_read_sqlite()` which takes a vector of table names 
or a connection and returns a named list of data frames (or if the `tibble` package is installed tibbles).
```{r}
tables <- rws_read_sqlite(conn)
names(tables)
tables$rws_data
```
It is worth noting that if the user wishes to read all the tables from a connection then they must pass the connection as the first argument even if the connection is specified in the options.

If the user wishes to read a single data frame they can use `rws_read_sqlite_table()`
```{r}
rws_read_sqlite_table("rws_data", conn = conn)
```

The `rws_read_sqlite_meta()` and `rws_read_sqlite_log()` allow the user to read the meta and log tables.
```{r}
rws_read_sqlite_meta(conn)
```

```{r}
rws_read_sqlite_log(conn)
```

## Cleaning Up

It's good practice to close a connection once you have finished with it.
```{r}
DBI::dbDisconnect(conn)
```
